Manuscript title

### 2018-12-03 Manuscript/manuscript.Rmd
manuscript edits

---
title: "Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings"
csl: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/styles.ref/genomebiology.csl
output: 
  word_document:
    #reference_docx: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/styles.doc/NIH_grant_style.docx
bibliography: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/References/references.bib
---

Spiro C. Stilianoudakis^1^ (stilianoudasc@mymail.vcu.edu), Mikhail G. Dozmorov^1\*^ (mikhail.dozmorov@vcuhealth.org)

^1^ Dept. of Biostatistics, Virginia Commonwealth University, Richmond, VA, 23298, USA  
^\*^ To whom correspondence should be addressed: Virginia Commonwealth University, Richmond, VA, 23298, 804-827-2055, mikhail.dozmorov@vcuhealth.org

@@

**Methods.** We propose a novel data pre-processing pipeline that addresses those shortcomings. Lasso and Elastic Net were used for feature selection, coupled with five classifier methods. A number of classifier performance metrics were assessed, including the F1 measure and Matthew Correlation Coefficient (MCC), and area under the ROC curve (AUROC).

**Results.** Data preprocessing (log2-transformation and standardization) improves the performance of classification algorithms and allows for the ability to more accurately predict which genomic features are most associated with TAD boundaries. In contrast to overlap-based association measures, the distance between genomic elements was the most predictive measure. The random undersampling strategy addresses the class imbalance problem the most effectively in nearly all settings.

**Conclusions.** Current methods used to model the epigenomic features associated with TAD boundaries are insufficiently robust to handle properties of genomic data. Models applied to unprocessed data can have poor predictive performances. Focusing solely on standard performance assessment metrics, such as AUROC, can mask poor performance of the models; thus, the use of more balanced metrics, such as F1 and MCC, is warranted. Our model results in better performances and more accurate identification of important features associated with the formation of TAD boundaries.

@@

We first compared models with either overlap counts or overlap percent predictor types, from the 5kb resolution data (Figure 12). These results were from random forest models fit to data without any re-sampling, using either LASSO or Elastic-Net regulariations. We found that differences in performances across variable reduction technique were marginal. However, when evaluating model performance using accuracy or AUC, the model appeared to perform relatively well. At second glance, though, when considering both the F1-score and MCC of the model, we were able to assess just how poorly the model was indeed performing (Supplementary Table 9). The low, and sometimes missing, values indicate that the model is unable to identify many, if any, true positive outcomes. That is, given an imbalanced data, the model is unable to predict which genomic bins truly contain a flanked TAD boundary. The performance results were found to be consistently poor for models with no re-sampling across the other resolutions (Supplementary Figure 13).

###Random UNDER-SAMPLING IMPROVES MODEL PERFORMANCE IN ALL RESOLUTIONS

Moving down the rows in each of the 4 regions of Figure 11 we see that model performance was markedly improved among the F1-scores and MCCs after creating balanced classes compared to the results using no re-sampling. For each of the resolutions, we see that random under-sampling out performed the other class balancing techniques. While SMOTE performed very similarly to RUS, ROS performed poorly in comparison, especially as the resolution of the data increased. Furthermore, we see that the elastic-net reduction technique outperformed the LASSO. Although, the disparity became less evident as the resolution increased.

>>>>

---
title: "Manuscript title"
#csl: styles.ref/genomebiology.csl
output: 
  word_document:
    #reference_docx: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/styles.doc/NIH_grant_style.docx
bibliography: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/References/references.bib
---

@@

**Methods.** We propose a novel data pre-processing pipeline that addresses those shortcomings. A number of classifier performance metrics were assessed, including the F1 measure and Matthew Correlation Coefficient (MCC), and area under the ROC curve (AUROC).

**Results.** Data preprocessing (log2-transformation and standardization) improves the performance of classification algorithms and allows for the ability to more accurately predict which genomic features are most associated with TAD boundaries.

**Conclusions.** Current methods used to model the epigenomic features associated with TAD boundaries are insufficiently robust to handle properties of genomic data. Models applied to unprocessed data can have poor predictive performances. Focusing solely on typical performance assessment metrics, such as AUROC, can mask poor performance of the models; thus, the use of more balanced metrics, such as F1 and MCC, is warranted. Our model results in better performances and more accurate identification of important features associated with the formation of TAD boundaries.

@@

We first compared models with either overlap counts or overlap percent predictor types, from the 5kb resolution data (Figure 12). These results were from random forest models fit to data without any re-sampling, using either LASSO or Elastic-Net regulariations. We found that differences in performances across variable reduction techniques were marginal. More importantly, when evaluating model performance using accuracy or AUC, the model appeared to perform relatively well. At second glance, however, when considering both the F1-score and MCC, we were able to assess just how poorly the model was performing (Supplementary Table 9).

The performances were found to be consistently poor for models with no re-sampling across the other resolutions (Supplementary Figure 13; Supplementary Table 10). The low, and sometimes missing, values indicate that the model is unable to identify many, if any, true positive outcomes. That is, given imbalanced data, the model was unable to predict which genomic bins truly contain a flanked TAD boundary.

###RANDOM UNDER-SAMPLING IMPROVES MODEL PERFORMANCE

When evaluatating the 5kb resolution data for a randomly under-sampled training set, it was found that models obtained better classification performance for F1 and MCC metrics (Figure 14). The increase in performance was present across both overlap predictor types. Models also featured a decrease in both accuracy and AUC, which was attributed to less inflated false negative rates (Supplementary Table 11). Differences in performances were again marginal across the two variable reduction techniques. Performances were also improved among the models using the other data resolutions. However, model improvement dissipated as resolution increased (Supplementary Figure 15; Supplementary Table 12).
